Experience shapes individual foraging specialization and success in a virtual predator-prey system

### 2023-10-26 manuscript/manuscript.Rmd
updated refs to have one file. updated in MS too

---
title: "Experience shapes individual foraging specialization and success in a virtual predator-prey system"
output: 
  word_document:
    reference_docx: manuscript-styles.docx
    fig_caption: yes
bibliography: 
  - references1.bib
  - references2.bib
csl: ecology.csl
---

@@

Predator express a diverse array of hunting tactics to capture their prey. In natural populations, differences among individuals in tactics such as ambushing, fast movement speeds, or high attack frequencies are associated with differences in prey consumption [@toscanoTraitmediatedFunctionalResponses2014;@Patrick.Weimerskirch2014;@toscanoPersonalityForagingBehavior2016]. Prey usually respond to predation cues by reducing their activity, or use rapid or unpredictable movements to successfully escape when they are chased [@Walker.etal2005;@Kelley.Magurran2011;@Herbert-Read.etal2017]. This results in reciprocal interactions where predators fine-tune their tactics to their prey, suggesting that they learn and become efficient hunters through repeated encounters [@Estes.etal2003;@Woo.etal2008;@Phillips.etal2017]. These behavioural adjustments with experience can shape the outcome of predator-prey interactions, which may result in larger implications for community and ecosystem dynamics [@wooster2023].

Empirical studies on human and non-human hunters show that individuals optimize foraging efficiency (e.g. search and handling times, return rates) by learning associative images or prey and environmental cues [@Edwards.Jackson1994;@Morse2000;@MacDonald2007;@Reid.etal2010;@Wilson-Rankin2015]. As predators practice and refine their tactics, they acquire expertise (e.g. skills, knowledge) that allows them to outperform novice hunters [@Dukas2019;@wooster2023]. Through this process, it is hypothesized that individuals will specialize in a hunting tactic (e.g. always moving fast) if it allows them to maintain successful prey captures at each atempt [@Tinker.etal2009;@Dukas2019]. When predators learn to hunt in environments where resources are predictable (or stable), they should gradually specialize in their hunting tactics with experience because individuals require less energy and time to search for, capture, and handle prey [@Weimerskirch2007;@Woo.etal2008;@Potier.etal2015]. These conditions favor specialist hunters because there would be higher costs to switch tactics for the same type of prey. An alternative hypothesis is that learning is a mechanism driving behavioural flexibility. For instance, predators can gain expertise and information on their prey through repeated hunting atempts, and may hunt optimally by learning to adjust their tactic to the type of prey that they encounter [@Stephens1993;@Ishii.Shimada2010;@Kelley.Magurran2011]. While there are costs to attempt novel tactics by trial and error in stable environments, predators that hunt in environments under fluctuating resources can reduce the consequences of uncertainty by learning to switch hunting tactics [@dall2010;@mathotAdaptiveStrategiesManaging2012]. Such feedbacks between ecological interactions and learning make it challenging to predict under which circumstances predators should specialize with experience [@Stephens1993;@Ishii.Shimada2010;@vilapouca2021;@beyts2023].

There is increasing evidence suggesting that the emergence of behavioural specialization and its fitness consequences is driven by resource fluctuations [@Patrick.Weimerskirch2014a;@vandenBosch.etal2019;@Santoro.etal2019;@Manlick.etal2021]. If predators that hunt in different environments adjust their tactics to local prey fluctuations, then specialist and flexible hunters can both attain similar hunting success and coexist [@Phillips.etal2017]. For example, @Szopa-Comley.Ioannou2022 showed that predatory blue acaras (*Andinoacara pulcher*) reduce their maximum approach speed when repeatedly exposed to prey that escape unpredictably, enabling them to be as successful as those exposed to predictable prey. Therefore, if specialist and flexible hunters can attain equal success by learning through repeated interactions with their prey, then empiricists should investigate the link between foraging specialization and success across time and contexts.

A recurring challenge impeding research on predator-prey behavioural interactions is the need to collect data simultaneously on both the predator and the prey. Individual-based models can mitigate some of these challenges and have contributed in stirring testable ideas for empiricists [cornell2019]. Yet, while providing maximum control over environmental parameters, they can suffer from emerging behaviours that are too complex to analyze mathematically [cornell2019]. Videogames offer an interesting alternative to agent-based simulations, providing controled virtual environments where real players interact. For example, in the predator-prey videogame *Dead by Daylight*, four prey players need to forage for resources while avoiding predation by a fifth player. In this virtual system, the predator population is composed of individuals that either ambush or hunt at high speeds (i.e. mean movement speed along a slow-fast continuum), and their success is driven by the movement of the prey [@FraserFranco.etal2022]. We found that prey can increase their chances of survival by cooperating and moving fast when they forage [@Cere.etal2021;@, Santostefano et al. in press]. Of course, the life of neither the predator nor the prey players are at stake, such that emerging patterns could be driven by their motivation to win and not "true" survival. Yet, the game ellicits natural reactions in players such as fear from predation (personal observations), which corroborates with another virtual ecological study showing that predation drives individual variation in risk perception [@Beauchamp2020]. In a different system, @lymberyetal.2023 showed that complex environments favor strong soldiers over large armies in a virtual game as well as in the laboratory with two ant species. These observations outline how ecological patterns can emerge from interactions in virtual systems with fixed rules. Videogames also generate large volumes of data on interacting players throughout their lifetime in the game under realistic, controlled, and repeatable ecological scenarios. Hence, virtual systems such as *Dead by Daylight* offer the opportunity to tackle fundamental questions about the role of experience on predator-prey interactions.

@@

*DBD* is a survival asymmetric multiplayer online game (i.e. a game where the gameplay mechanics differ between two groups) developed by Behaviour Interactive Inc, in which players can play either as a predator or a prey. Every match includes only one predator and four prey. The objective of the predator is to hunt and capture the prey, and the objective of the prey is to search for resources while avoiding the predator. The resources are in the form of power generators that, once all activated, will enable the prey to escape through one of two exit doors. The composition of the predator and prey group in a match is determined by a skill-based matchmaking algorithm. A match ends when the predator kills all the prey available (i.e. that have not escaped), or when the last remaining prey escapes the virtual environment. Each player, predator or prey, can choose an avatar with abilities that encourage specific play styles (e.g. bold vs cautious prey, or ambush vs roaming predator). During our study period, the game offered `r toString(format(length(unique(dat$avatar_id))))` predator avatars. The virtual environments are composed of fixed and procedurally generated habitat components, such as vegetation, mazes, and buildings. Some of these environments are larger than others, with varying structural complexity. However, predators display only minimal changes in behaviour and hunting success across the environments, probably due to a game feature enabling them to have visual cues of the generators [@FraserFranco.etal2022]. There were `r toString(format(length(unique(dat$environment_id))))` virtual game environments available for play during the study period.

@@

All models were fitted under a Bayesian framework in R (version 4.1.2) using Markov chain Monte Carlo (MCMC) sampling with the package "brms" version 2.16.3 [@Burkner2017], an R front-end for the STAN software [@StanDevelopmentTeam2023], and "cmdstanr" version 0.4.0 [@Gabry.Cesnovar2021] as the back-end for estimation (cmdstan installation version 2.28.2). The models were run on Cedar (Operating system: CentOS Linux 7), a computer cluster maintained by the Digital Research Alliance of Canada (https://docs.alliancecan.ca/wiki/Cedar). Each required 64GB of RAM with 48 cores to compile within 5 days.

@@

We tested how prey influenced the development of predator expertise by computing five Bayesian generalized additive mixed models (GAMM) with thin-plate regression splines, all of which estimated the relationship between hunting success (i.e. number of prey consumed) and the predators' cumulative experience (i.e. number of matches played before the current match). We parametrized the models following the method of @Pedersen.etal2019. The first model was the simplest, with a common global smoothing function and random intercepts for the predator ID. In this model, we assume that predators have the same development of expertise, with the model estimating a trend for the average individual (i.e. global smoother). For the second model, we included varying individual smoothers for the predator ID. Here, we assume that individual predators share a similar relationship between success and experience, but that this relationship can vary among them (e.g. predator 1 has a steeper curve than predator 2). This enabled us to test whether predators differed in the development of their expertise. In the third model, we kept the individual smoothers for the predators, but removed the global smoother. This model assumes that predators do not share a common relationship between success and experience. The fourth and fifth models were reproductions of the second and third models respectively, where we included the prey speed to assess its effect on the relationship between success and experience. We included the standardized match duration and prey rank as covariates in all five models.

We computed the five models using a modified version of the beta-binomial distribution implemented in "brms". Hunting success was estimated as the probability of consuming the four prey ($\mu_i$), drawn from a Beta distribution ($Beta(\mu_i, \phi)$) with mean ($\mu \in [0,1]$) and precision ($\phi > 0$) parameters. We used a logit link function to estimate $\mu_i$ where $\mu_i = \frac{exp(\eta_i)}{exp(\eta_i) + 1}$ and $\eta_i$ is the linear predictor, while the precision parameter ($\phi$) was estimated with an identity link. We used the default number of basis functions (K) in "brms" for the models to estimate the relationship between hunting success and experience. We assumed that the random intercepts for the predator ID ($id$) followed a Gaussian distribution with estimated standard deviation ($id \sim N(0,\ I\ \sigma_{id}^{2})$). We used weakly informative Gaussian priors for the intercept ($N(0,\ 0.5^{2})$) and the global trend of cumulative experience ($N(0,\ 2^{2}))$). Following @FraserFranco.etal2022, we defined a positive Gaussian prior on the precision parameter ($N(2,\ 0.5^{2})$), a positive Gaussian prior ($N(1,\ 0.5^{2}))$) on the game duration because longer trials lead to greater success, and a negative Gaussian prior on prey speed ($N(-1,\ 0.5^{2})$) because encountering faster prey is associated with lower success in this system. We employed weakly informative half-Gaussian priors on all the standard deviation parameters ($N(0,\ 0.5^{2})$). We compared the models' predictive accuracies using approximate leave-one-out cross-validation with Pareto-smoothed importance sampling [@Vehtari.etal2017;@Piironen.Vehtari2017;@Vehtari.etal2022].

@@

We tested whether foraging behaviour changed with experience, and compared the success of predators along the flexible-specialist hunter continuum, using a multivariate double-hierarchical generalized linear model (MDHGLM) [@Lee.Nelder2006;@Cleasby.etal2015;@Mitchell.etal2016a]. For each experience level, we fitted the model on three traits: the predator's speed ($Y_{1}$) and the prey's speed ($Y_{2}$) using a Gaussian error distribution, and the predator's hunting success ($Y_{3}$) using a beta-binomial error distribution. Fixed and random effect structure were trait-specific and the same for all experience levels. The predator and the prey speed included a mean and a dispersion part (eqn 1.1, 1.2 and 2.1, 2.2). For both traits, we controlled for the standardized prey rank on both the mean and the dispersion parts of the model. We included game environment, predator avatar, and individual predator identity random effects on the mean part of the model. The dispersion part included individual predator identity random effects exclusively. For hunting success, we modeled the mean part only (eqn 3) and controlled for the standardized prey rank and game duration, and included an individual identity random effect.

@@

We used weakly informative Gaussian priors for the prey rank ($N(0,\ 1^{2})$: eqn. 1.1 to 2.2) and game duration ($N(0.5,\ 0.5^{2})$: eqn. 3). Based on previous results showing that the mean predator speed revolves around 3 m/s [@FraserFranco.etal2022], we applied a weakly informative Gaussian prior on the intercept for predator and prey speed ($N(3,\ 0.5^{2})$: eqn. 1.1 and 2.1). Similarly, we used a weakly informative Gaussian prior on the intercept for the dispersion part ($N(0,\ 1^{2})$: eqn 1.2 and 2.2). For hunting success, we applied a positive Gaussian prior on the precision parameter ($N(2,\ 1^{2})$: eqn. 3) and a negative Gaussian prior ($N(-4,\ 1^{2})$: eqn 3) on the intercept because hunting success should be close to 0 at the lowest game duration. For every predicted variable, we applied a weakly informative half-Gaussian prior ($N(0,\ 1^{2})$) on the random effects for the mean part and the dispersion part of the model. We used a regularizing $LKJ(2)$ prior on the correlation matrix.

@@

We parametrized the GAMMs and the MDHGLM to run four MCMC chains with 1000 posterior samples for each parameter. We ran 2500 iterations with a thinning set to eight for the additive model with a global smoother only (see Table 1), and 1500 iterations with a thinning set to four for the other additive models. We set the first 500 iterations in each model as warm ups. We parametrized the MDHGLM to run 2500 iterations with a thinning set to eight, with the first 500 iterations used as warm ups. We assessed the convergence of the MCMC chains using trace plots, R-hat diagnostics with a threshold of <1.01, and effective sample sizes (ESS) with a threshold of >100 [@Vehtari.etal2021a]. We also performed posterior predictive checks which showed an adequate fit of the models (for details, see https://github.com/quantitative-ecologist/experience-hunting-tactics). We report all the posterior parameter estimates using the median of the posterior distribution with the highest posterior density (HPD) intervals at 50%, 80%, and 95%.

@@

Learning from experience is predicted to increase individual predator foraging specialization when prey distribution, behaviour, or movement is predictable through time [@Estes.etal2003;@Tinker.etal2008;@Tinker.etal2009]. However, if predators experience unpredictable encounters with their prey, then learning should increase behavioural flexibility [@Stephens1993;@Ishii.Shimada2010]. We found that predators in *Dead by Daylight* increased their hunting success with experience, suggesting that predator expertise was honed through extensive practice. While there was no directional increase in either behavioural specialization or flexibility with experience at the population level, individual predators developed their own behavioural trajectories over successive encounters with the prey. This was characterized by a continuum of hunting styles ranging from a slower and flexible strategy to a highly specialized fast-paced strategy, with some predators switching between the two with experience. The behavioural strategies of predators and prey tended to match across experience, suggesting that both were adjusting to each other. Specialized cursorial hunters encountered similar groups of prey, while slower and flexible hunters encountered varying groups of prey. Overall, predators along the flexible-specialist continuum achieved similar success, although flexible hunters were slightly less successful as advanced hunters.

@@

Our results suggest that predator expertise is honed through extensive practice. The predator population increased and stabilized its success with experience in a diminishing returns fashion as it is typically found in empirical studies of expertise [reviewed in @Dukas2019]. The prey were important in mediating this pattern, probably because they increased their speed as predators gained experience. We previously showed in *DBD* that faster movement is an effective strategy used by the prey to avoid predation [@FraserFranco.etal2022], and other studies have found that as well [@Walker.etal2005;@Kelley.Magurran2011;@Martin.etal2022]. This resulted in discernable differences in the relationship between success and experience among predators, implying that the development of expertise depends on an individual's capacity to adjust to difficulty. Hunting faster prey requires costly and specialized cognitive abilities and coordination in predators [@Kelley.Magurran2011], and that could explain why such prey were more difficult to capture. Thus, predators that couldn't properly hunt at high speeds or develop counter-strategies for faster prey were likely at a disadvantage.

However, even after we took the prey's speed into account, there were still important differences in expertise acquisition among predators, suggesting that other antipredator tactics were potentially involved. For example, if the prey used camouflage or hiding to avoid being detected [@Bowen.etal2002;@Carey.Wahl2011;@Nomura.etal2011], then predators may have found them only if they learned how to exploit the visual cues that facilitate cryptic prey detection, such as habitat characteristics or prey colour patterns [@Ehlinger1989;@Hughes.etal1992;@Warburton2003;@Szopa-Comley.etal2020]. In addition, longer time intervals between hunting events may delay or even impede learning due to forgetting important information [@Endler1991;@Wright.etal2022]. In *DBD*, a predator that played 300 matches in the span of six months might forget more critical information related to prey detection or escape patterns than one that played 300 matches in the span of six days. Investigating the impact of such time lags in future analyses may reveal important insights on the outcome of predator-prey interactions.

@@

The predator population maintained its speed and degree of specialization throughout the study period, and individuals differed only slightly in their average speeds. However, differences in specialization among individuals were important and increased slightly with experience. Predators hunting at higher speeds specialized in this tactic and tended to encounter groups of prey with similar speeds, while those hunting at slower speeds were flexible and tended to encounter varying groups of prey. On one hand, this pattern could arise if some predators adjusted their strategy to their prey, because fast-paced hunting is a specialized tactic suited for prey that use rapid evasive movements, while a slower and flexible tactic can be useful to minimize the consequences of uncertainty when prey escape unpredictably [@Endler1991;@Bro-Jorgensen2013;@Wilson.etal2018;@Szopa-Comley.Ioannou2022]. However, we observed that 44% of the predator population showed close to no change in specialization with experience. Since these individuals were already specialized as novices (Fraser Franco, personal observation), there would be no point in them changing their tactic with experience if they were already successful, or there could have been costs to switch their strategy if they more often encountered faster prey. This could explain why the behavioural interactions between the predator and the prey remained stable across the predator's experience. Yet, because prey can also learn how to avoid predation [@kelley2003;@turner2006;@mccomb2011], we believe that the predator-prey phenotype matching more likely emerged from reciprocal adjustments in speed by predators and prey as they interacted [@Kishida.etal2006;@Kishida.etal2009;@Edgell.Rochette2009a;@McGhee.etal2013a]. Indeed, 56% of the predator population altered its behaviour with experience to different degrees, including 12% that even switched between flexibility and specialization. Hence, if the prey also learned through repeated interactions with the predators, it is possible that experience contributed in stabilizing the system as both were adjusting to each other, similar to Red Queen dynamics [@brockhurst2014]. Our results appear to support this since many individuals were shifting their behavioural niche while the population niche remained stable across experience.

>>>>

---
title: "Experience shapes individual foraging specialization and success in a virtual predator-prey system"
output: 
  word_document:
    reference_docx: manuscript-styles.docx
    fig_caption: yes
bibliography: references.bib
csl: ecology.csl
---

@@

Predator express a diverse array of hunting tactics to capture their prey. In natural populations, differences among individuals in tactics such as ambushing, fast movement speeds, or high attack frequencies are associated with differences in prey consumption [@toscano2014;@patrick.weimerskirch2014a;@toscano.etal2016]. Prey usually respond to predation cues by reducing their activity, or use rapid or unpredictable movements to successfully escape when they are chased [@walker.etal2005;@kelley.magurran2011;@herbert-read.etal2017]. This results in reciprocal interactions where predators fine-tune their tactics to their prey, suggesting that they learn and become efficient hunters through repeated encounters [@estes.etal2003;@woo.etal2008;@phillips.etal2017]. These behavioural adjustments with experience can shape the outcome of predator-prey interactions, which may result in larger implications for community and ecosystem dynamics [@wooster.etal2023].

Empirical studies on human and non-human hunters show that individuals optimize foraging efficiency (e.g. search and handling times, return rates) by learning associative images or prey and environmental cues [@edwards.jackson1994;@morse2000;@macdonald2007;@reid.etal2010;@wilson-rankin2015]. As predators practice and refine their tactics, they acquire expertise (e.g. skills, knowledge) that allows them to outperform novice hunters [@dukas2019;@wooster.etal2023]. Through this process, it is hypothesized that individuals will specialize in a hunting tactic (e.g. always moving fast) if it allows them to maintain successful prey captures at each atempt [@tinker.etal2009;@dukas2019]. When predators learn to hunt in environments where resources are predictable (or stable), they should gradually specialize in their hunting tactics with experience because individuals require less energy and time to search for, capture, and handle prey [@weimerskirch2007;@woo.etal2008;@potier.etal2015]. These conditions favor specialist hunters because there would be higher costs to switch tactics for the same type of prey. An alternative hypothesis is that learning is a mechanism driving behavioural flexibility. For instance, predators can gain expertise and information on their prey through repeated hunting atempts, and may hunt optimally by learning to adjust their tactic to the type of prey that they encounter [@stephens1993;@ishii.shimada2010;@kelley.magurran2011]. While there are costs to attempt novel tactics by trial and error in stable environments, predators that hunt in environments under fluctuating resources can reduce the consequences of uncertainty by learning to switch hunting tactics [@dall2010;@mathot.etal2012]. Such feedbacks between ecological interactions and learning make it challenging to predict under which circumstances predators should specialize with experience [@stephens1993;@ishii.shimada2010;@vilapouca2021;@beyts2023].

There is increasing evidence suggesting that the emergence of behavioural specialization and its fitness consequences is driven by resource fluctuations [@patrick.weimerskirch2014b;@vandenbosch.etal2019;@santoro.etal2019;@manlick.etal2021]. If predators that hunt in different environments adjust their tactics to local prey fluctuations, then specialist and flexible hunters can both attain similar hunting success and coexist [@phillips.etal2017]. For example, @szopa-comley.ioannou2022 showed that predatory blue acaras (*Andinoacara pulcher*) reduce their maximum approach speed when repeatedly exposed to prey that escape unpredictably, enabling them to be as successful as those exposed to predictable prey. Therefore, if specialist and flexible hunters can attain equal success by learning through repeated interactions with their prey, then empiricists should investigate the link between foraging specialization and success across time and contexts.

A recurring challenge impeding research on predator-prey behavioural interactions is the need to collect data simultaneously on both the predator and the prey. Individual-based models can mitigate some of these challenges and have contributed in stirring testable ideas for empiricists [@cornell2019]. Yet, while providing maximum control over environmental parameters, they can suffer from emerging behaviours that are too complex to analyze mathematically [@cornell2019]. Videogames offer an interesting alternative to agent-based simulations, providing controled virtual environments where real players interact. For example, in the predator-prey videogame *Dead by Daylight*, four prey players need to forage for resources while avoiding predation by a fifth player. In this virtual system, the predator population is composed of individuals that either ambush or hunt at high speeds (i.e. mean movement speed along a slow-fast continuum), and their success is driven by the movement of the prey [@fraserfranco.etal2022]. We found that prey can increase their chances of survival by cooperating and moving fast when they forage [@cere.etal2021;@, Santostefano et al. in press]. Of course, the life of neither the predator nor the prey players are at stake, such that emerging patterns could be driven by their motivation to win and not "true" survival. Yet, the game ellicits natural reactions in players such as fear from predation (personal observations), which corroborates with another virtual ecological study showing that predation drives individual variation in risk perception [@beauchamp2020]. In a different system, @lymbery.etal2023 showed that complex environments favor strong soldiers over large armies in a virtual game as well as in the laboratory with two ant species. These observations outline how ecological patterns can emerge from interactions in virtual systems with fixed rules. Videogames also generate large volumes of data on interacting players throughout their lifetime in the game under realistic, controlled, and repeatable ecological scenarios. Hence, virtual systems such as *Dead by Daylight* offer the opportunity to tackle fundamental questions about the role of experience on predator-prey interactions.

@@

*DBD* is a survival asymmetric multiplayer online game (i.e. a game where the gameplay mechanics differ between two groups) developed by Behaviour Interactive Inc, in which players can play either as a predator or a prey. Every match includes only one predator and four prey. The objective of the predator is to hunt and capture the prey, and the objective of the prey is to search for resources while avoiding the predator. The resources are in the form of power generators that, once all activated, will enable the prey to escape through one of two exit doors. The composition of the predator and prey group in a match is determined by a skill-based matchmaking algorithm. A match ends when the predator kills all the prey available (i.e. that have not escaped), or when the last remaining prey escapes the virtual environment. Each player, predator or prey, can choose an avatar with abilities that encourage specific play styles (e.g. bold vs cautious prey, or ambush vs roaming predator). During our study period, the game offered `r toString(format(length(unique(dat$avatar_id))))` predator avatars. The virtual environments are composed of fixed and procedurally generated habitat components, such as vegetation, mazes, and buildings. Some of these environments are larger than others, with varying structural complexity. However, predators display only minimal changes in behaviour and hunting success across the environments, probably due to a game feature enabling them to have visual cues of the generators [@fraserfranco.etal2022]. There were `r toString(format(length(unique(dat$environment_id))))` virtual game environments available for play during the study period.

@@

All models were fitted under a Bayesian framework in R (version 4.1.2) using Markov chain Monte Carlo (MCMC) sampling with the package "brms" version 2.16.3 [@burkner2017], an R front-end for the STAN software [@StanDevelopmentTeam2023], and "cmdstanr" version 0.4.0 [@gabry.cesnovar2021] as the back-end for estimation (cmdstan installation version 2.28.2). The models were run on Cedar (Operating system: CentOS Linux 7), a computer cluster maintained by the Digital Research Alliance of Canada (https://docs.alliancecan.ca/wiki/Cedar). Each required 64GB of RAM with 48 cores to compile within 5 days.

@@

We tested how prey influenced the development of predator expertise by computing five Bayesian generalized additive mixed models (GAMM) with thin-plate regression splines, all of which estimated the relationship between hunting success (i.e. number of prey consumed) and the predators' cumulative experience (i.e. number of matches played before the current match). We parametrized the models following the method of @pedersen.etal2019. The first model was the simplest, with a common global smoothing function and random intercepts for the predator ID. In this model, we assume that predators have the same development of expertise, with the model estimating a trend for the average individual (i.e. global smoother). For the second model, we included varying individual smoothers for the predator ID. Here, we assume that individual predators share a similar relationship between success and experience, but that this relationship can vary among them (e.g. predator 1 has a steeper curve than predator 2). This enabled us to test whether predators differed in the development of their expertise. In the third model, we kept the individual smoothers for the predators, but removed the global smoother. This model assumes that predators do not share a common relationship between success and experience. The fourth and fifth models were reproductions of the second and third models respectively, where we included the prey speed to assess its effect on the relationship between success and experience. We included the standardized match duration and prey rank as covariates in all five models.

We computed the five models using a modified version of the beta-binomial distribution implemented in "brms". Hunting success was estimated as the probability of consuming the four prey ($\mu_i$), drawn from a Beta distribution ($Beta(\mu_i, \phi)$) with mean ($\mu \in [0,1]$) and precision ($\phi > 0$) parameters. We used a logit link function to estimate $\mu_i$ where $\mu_i = \frac{exp(\eta_i)}{exp(\eta_i) + 1}$ and $\eta_i$ is the linear predictor, while the precision parameter ($\phi$) was estimated with an identity link. We used the default number of basis functions (K) in "brms" for the models to estimate the relationship between hunting success and experience. We assumed that the random intercepts for the predator ID ($id$) followed a Gaussian distribution with estimated standard deviation ($id \sim N(0,\ I\ \sigma_{id}^{2})$). We used weakly informative Gaussian priors for the intercept ($N(0,\ 0.5^{2})$) and the global trend of cumulative experience ($N(0,\ 2^{2}))$). Following @fraserfranco.etal2022, we defined a positive Gaussian prior on the precision parameter ($N(2,\ 0.5^{2})$), a positive Gaussian prior ($N(1,\ 0.5^{2}))$) on the game duration because longer trials lead to greater success, and a negative Gaussian prior on prey speed ($N(-1,\ 0.5^{2})$) because encountering faster prey is associated with lower success in this system. We employed weakly informative half-Gaussian priors on all the standard deviation parameters ($N(0,\ 0.5^{2})$). We compared the models' predictive accuracies using approximate leave-one-out cross-validation with Pareto-smoothed importance sampling [@vehtari.etal2017;@piironen.vehtari2017;@vehtari.etal2022].

@@

We tested whether foraging behaviour changed with experience, and compared the success of predators along the flexible-specialist hunter continuum, using a multivariate double-hierarchical generalized linear model (MDHGLM) [@lee.nelder2006;@cleasby.etal2015;@mitchell.etal2016]. For each experience level, we fitted the model on three traits: the predator's speed ($Y_{1}$) and the prey's speed ($Y_{2}$) using a Gaussian error distribution, and the predator's hunting success ($Y_{3}$) using a beta-binomial error distribution. Fixed and random effect structure were trait-specific and the same for all experience levels. The predator and the prey speed included a mean and a dispersion part (eqn 1.1, 1.2 and 2.1, 2.2). For both traits, we controlled for the standardized prey rank on both the mean and the dispersion parts of the model. We included game environment, predator avatar, and individual predator identity random effects on the mean part of the model. The dispersion part included individual predator identity random effects exclusively. For hunting success, we modeled the mean part only (eqn 3) and controlled for the standardized prey rank and game duration, and included an individual identity random effect.

@@

We used weakly informative Gaussian priors for the prey rank ($N(0,\ 1^{2})$: eqn. 1.1 to 2.2) and game duration ($N(0.5,\ 0.5^{2})$: eqn. 3). Based on previous results showing that the mean predator speed revolves around 3 m/s [@fraserfranco.etal2022], we applied a weakly informative Gaussian prior on the intercept for predator and prey speed ($N(3,\ 0.5^{2})$: eqn. 1.1 and 2.1). Similarly, we used a weakly informative Gaussian prior on the intercept for the dispersion part ($N(0,\ 1^{2})$: eqn 1.2 and 2.2). For hunting success, we applied a positive Gaussian prior on the precision parameter ($N(2,\ 1^{2})$: eqn. 3) and a negative Gaussian prior ($N(-4,\ 1^{2})$: eqn 3) on the intercept because hunting success should be close to 0 at the lowest game duration. For every predicted variable, we applied a weakly informative half-Gaussian prior ($N(0,\ 1^{2})$) on the random effects for the mean part and the dispersion part of the model. We used a regularizing $LKJ(2)$ prior on the correlation matrix.

@@

We parametrized the GAMMs and the MDHGLM to run four MCMC chains with 1000 posterior samples for each parameter. We ran 2500 iterations with a thinning set to eight for the additive model with a global smoother only (see Table 1), and 1500 iterations with a thinning set to four for the other additive models. We set the first 500 iterations in each model as warm ups. We parametrized the MDHGLM to run 2500 iterations with a thinning set to eight, with the first 500 iterations used as warm ups. We assessed the convergence of the MCMC chains using trace plots, R-hat diagnostics with a threshold of <1.01, and effective sample sizes (ESS) with a threshold of >100 [@vehtari.etal2021]. We also performed posterior predictive checks which showed an adequate fit of the models (for details, see https://github.com/quantitative-ecologist/experience-hunting-tactics). We report all the posterior parameter estimates using the median of the posterior distribution with the highest posterior density (HPD) intervals at 50%, 80%, and 95%.

@@

Learning from experience is predicted to increase individual predator foraging specialization when prey distribution, behaviour, or movement is predictable through time [@estes.etal2003;@tinker.etal2008;@tinker.etal2009]. However, if predators experience unpredictable encounters with their prey, then learning should increase behavioural flexibility [@stephens1993;@ishii.shimada2010]. We found that predators in *Dead by Daylight* increased their hunting success with experience, suggesting that predator expertise was honed through extensive practice. While there was no directional increase in either behavioural specialization or flexibility with experience at the population level, individual predators developed their own behavioural trajectories over successive encounters with the prey. This was characterized by a continuum of hunting styles ranging from a slower and flexible strategy to a highly specialized fast-paced strategy, with some predators switching between the two with experience. The behavioural strategies of predators and prey tended to match across experience, suggesting that both were adjusting to each other. Specialized cursorial hunters encountered similar groups of prey, while slower and flexible hunters encountered varying groups of prey. Overall, predators along the flexible-specialist continuum achieved similar success, although flexible hunters were slightly less successful as advanced hunters.

@@

Our results suggest that predator expertise is honed through extensive practice. The predator population increased and stabilized its success with experience in a diminishing returns fashion as it is typically found in empirical studies of expertise [reviewed in @dukas2019]. The prey were important in mediating this pattern, probably because they increased their speed as predators gained experience. We previously showed in *DBD* that faster movement is an effective strategy used by the prey to avoid predation [@fraserfranco.etal2022], and other studies have found that as well [@walker.etal2005;@kelley.magurran2011;@martin.etal2022]. This resulted in discernable differences in the relationship between success and experience among predators, implying that the development of expertise depends on an individual's capacity to adjust to difficulty. Hunting faster prey requires costly and specialized cognitive abilities and coordination in predators [@kelley.magurran2011], and that could explain why such prey were more difficult to capture. Thus, predators that couldn't properly hunt at high speeds or develop counter-strategies for faster prey were likely at a disadvantage.

However, even after we took the prey's speed into account, there were still important differences in expertise acquisition among predators, suggesting that other antipredator tactics were potentially involved. For example, if the prey used camouflage or hiding to avoid being detected, then predators may have found them only if they learned how to exploit the visual cues that facilitate cryptic prey detection, such as habitat characteristics or prey colour patterns [@ehlinger1989;@hughes.etal1992;@warburton2003;@szopa-comley.etal2020]. In addition, longer time intervals between hunting events may delay or even impede learning due to forgetting important information [@endler1991;@wright.etal2022]. In *DBD*, a predator that played 300 matches in the span of six months might forget more critical information related to prey detection or escape patterns than one that played 300 matches in the span of six days. Investigating the impact of such time lags in future analyses may reveal important insights on the outcome of predator-prey interactions.

@@

The predator population maintained its speed and degree of specialization throughout the study period, and individuals differed only slightly in their average speeds. However, differences in specialization among individuals were important and increased slightly with experience. Predators hunting at higher speeds specialized in this tactic and tended to encounter groups of prey with similar speeds, while those hunting at slower speeds were flexible and tended to encounter varying groups of prey. On one hand, this pattern could arise if some predators adjusted their strategy to their prey, because fast-paced hunting is a specialized tactic suited for prey that use rapid evasive movements, while a slower and flexible tactic can be useful to minimize the consequences of uncertainty when prey escape unpredictably [@endler1991;@bro-jorgensen2013;@wilson.etal2018;@szopa-comley.ioannou2022]. However, we observed that 44% of the predator population showed close to no change in specialization with experience. Since these individuals were already specialized as novices (Fraser Franco, personal observation), there would be no point in them changing their tactic with experience if they were already successful, or there could have been costs to switch their strategy if they more often encountered faster prey. This could explain why the behavioural interactions between the predator and the prey remained stable across the predator's experience. Yet, because prey can also learn how to avoid predation [@kelley2003;@turner2006;@mccomb2011], we believe that the predator-prey phenotype matching more likely emerged from reciprocal adjustments in speed by predators and prey as they interacted [@kishida.etal2006;@kishida.etal2009;@edgell.rochette2009;@mcghee.etal2013]. Indeed, 56% of the predator population altered its behaviour with experience to different degrees, including 12% that even switched between flexibility and specialization. Hence, if the prey also learned through repeated interactions with the predators, it is possible that experience contributed in stabilizing the system as both were adjusting to each other, similar to Red Queen dynamics [@brockhurst2014]. Our results appear to support this since many individuals were shifting their behavioural niche while the population niche remained stable across experience.

Classical theory predicts that individual foraging specialization should maximize prey consumption by optimizing foraging efficiency [@stephens1986], but other work suggests that flexibility and specialization could both be adaptive as resources fluctuate [@woo.etal2008;@phillips.etal2017]. Our results support empirical findings indicating that flexible and specialist foragers obtained similar success [@weimerskirch2007;@woo.etal2008;@potier.etal2015;@phillips.etal2017;@courbin.etal2018], although flexible foragers were slightly less successful. Both strategies appeared to emerge in part from individuals learning how to hunt their prey, resulting in a general increase in success in the population. However, there were still considerable differences in success among individuals through time, suggesting that some predators were limited in their capacity to match their tactic to their prey or to increase their success. Specialist foragers were faster, and thus, probably better equipped to hunt the more difficult faster prey in *DBD*. However, if the prey responded to fast predators by also being faster, then hunting at high speeds resulted in more difficult encounters for the predator, thereby decreasing the benefits of using this tactic (Figure 4). Thus, specializing probably compensated for the difficulty of hunting prey at high speeds by helping predators to better predict the location and movement of their prey. On the other hand, flexible foragers encountered variable prey with slower speeds. Yet, because the prey increased their speed with experience, the benefits of being able to hunt multiple prey types for flexible hunters may have come at the cost of not being adept at capturing faster prey [@pintor.etal2014]. Such trade-offs may reflect limitations in learning all the skills required to successfully hunt all types of prey [@healy1992;@belisle1997;@dukas2019], particularly if the skills required to hunt slower prey are nontransferable to faster prey. As the benefits of each hunting style changed under different scenarios, the combination of predatory styles in our virtual system reflects how resource fluctuations can maintain fitness equilibrium within populations, resulting in behavioural diversity in predator-prey systems [@tinker.etal2008;@woo.etal2008;@mougi.kishida2009].
